Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments

### 2020-04-10 peakcharacterization_manuscript.Rmd
added conversion back into M/Z into manuscript

@@

Scans are normalized to a single *median* scan based on the log-intensity differences of a subset of peaks that are present in at least the same number of scans as the 95th percentile of scan counts for the peaks. 
In addition, only those peaks that have an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans. 
The scan with the lowest summed overall distance is chosen as the "reference" scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan. 
This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan number is checked, and those peaks with correlation of greater than 0.5 with scan number are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.

@@

High-peak-density (HPD) artifacts [Mitchell et al., 2018] present as singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations. Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and those FSD that are greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work.
For HPD detection, the peaks in excel output from Xcalibur were used after converting the M/Z peak centers into frequency space. Sliding windows 1000 frequency points wide and offset by 100 points were used for the density calculations.

@@

We previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [Mitchell et al., Metabolomics, 2018]. 
Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks. 
Their presence should be minimized by removing noise peaks, and removing regions that have multiple reported peaks in the same scan. 
We expect they may present are as characterized peaks that have larger than expected standard deviations in their frequencies when calculated across scans.
These peaks can be detected by simply examining the distribution of frequency SDs and removing those that are outliers.
To verify the removal of HPD regions, we converted centroided M/Z's from XCalibur to frequency values using the previously decided values for that sample.
Peak density was measured using a sliding window 10 points wide and offset by one point.

@@

As an alternative to RSD, we can also compare the fit of relative intensities after assignment using SMIRFE [reference] to the theoretical relative natural abundances (relNAP) of the assigned isotopic molecular formula's (IMFs) within the assigned elemental molecular formula's (EMFs).
Theoretically, we expect that lower quality data will have both lower numbers of assignments, and that for those things that are assigned, the fit between relative intensity and relNAP will be worse.

@@

For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class. 
If incorrect peak intensities are being reported, then we would expect between class to within class variances to be very random, resulting in F-statistics close to 1.
For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using our peak-characterization and those from XCMS.

>>>>

To convert M/Z back into frequency, we can use a similar model without the roots.

$$mz = intercept + x* \frac{1}{frequency} + y*\frac{1}{frequency^2} + z*\frac{1}{frequency^3}$$

@@

Scans are normalized to a single *reference* scan based on the log-intensity differences of a subset of peaks that are present in at least the same number of scans as the 95th percentile of scan counts for the peaks. 
In addition, only those peaks that have an intensity greater than 0.7 times the highest intensity peak in the scan are used. 
Pairwise scan-to-scan distances are calculated by taking the cartesian distances between log peak intensities present in both scans, and then the cartesian distance is summed across the scan-to-scan distance to provide an overall difference of each scan to all other scans. 
The scan with the lowest summed overall distance is chosen as the *reference* scan, and normalization factors for each scan are calculated as the median log peak intensity differences in $scan_i$ compared to the reference scan. 
This normalization is done twice, once using all possible peaks, after which the correlation of peak intensity with scan number is checked, and those peaks with correlation of greater than 0.5 with scan number are removed, and the normalization factors are calculated again, and then applied to both the centroided peak height and the raw point intensities.

@@

High-peak-density (HPD) artifacts [Mitchell et al., 2018] present as singular peaks with higher than expected frequency standard deviations (FSDs) calculated from the scan-to-scan frequency peak locations. Outliers are detected by calculating the interquartile range (IQR) of the distribution of FSDs across the entire spectrum, and those FSD that are greater than the median plus 1.5 times the IQR (as implemented in boxplot.stats) are removed.
The HPD detection algorithm from Mitchell et al. was re-implemented in R for this work to allow comparisons between it and the use of the FSD.
For HPD detection, the peaks in excel output from Xcalibur were used after converting the M/Z peak centers into frequency space. Sliding windows 1000 frequency points wide and offset by 100 points were used for the density calculations.

### Multi-Class F-Statistics

To evaluate the F-statistic between multi-class samples, a much larger set of samples was used. In this case, 181 matched non-cancer and cancer samples that were previously used in the HPD detection manuscript [ref]. All 181 samples were peak characterized using the most stringent method below as well as peaks generated from the xcms R package [ref], assigned using SMIRFE, and then peaks matched by shared EMFs across samples (see Supplemental materials for a description of peak matching and deciding the most likely EMF). For each matched peak across samples, we kept the peaks present in at least 25% of samples in both classes of sample (disease, non-disease), and calculated the F-statistic for peaks generated using our peak-characterization and those from xcms.

@@

We have previously described the presence of high peak density (HPD) artefacts in FT-MS spectra [Mitchell et al., Metabolomics, 2018]. 
Ideally, the peak characterization procedure should reduce their presence in the resultant reported peaks. 
Their presence should be minimized by removing noise peaks, and removing regions that have multiple reported peaks in the same scan. 
We expect they may present as characterized peaks that have larger than expected frequency standard deviations (FSD) when calculated across scans.
These peaks can be detected by simply examining the distribution of frequency SDs and removing those that are outliers.
To verify the removal of HPD regions, we converted centroided M/Z's from XCalibur to frequency values using the previously calculated values for that sample.
Peak density was measured using a sliding window 10 points wide and offset by one point.

Figure X compares the HPD regions detected using the method from Mitchell et al., and their correspondence with high FSD peaks for one sample (the other comparisons are in Supplemental materials).
Based on this comparison, the high FSD peaks are indeed HPD sites and should be removed before assignment.
However, the FSD does not depend on calculating a sliding window based density, but is merely a product of those peaks across multiple scans that span a wider range frequency range than expected based on the peaks in the data. We also note that we do expect higher M/Z standard deviations at higher M/Z, but the frequency based method does not suffer the same effect, due to the offsets of frequency points being constant.

@@

As an alternative to RSD, we can also compare the fit of relative intensities after assignment using SMIRFE [reference] to the theoretical relative natural abundances (relNAP) of the assigned isotopic molecular formula's (IMFs) within the assigned elemental molecular formula's (EMFs).
Theoretically, we expect that lower quality data will have both lower numbers of assignments, and that for those things that are assigned, the fit between relative intensity and relNAP will be worse.
To compare relative NAP to relative abundances, we only examined the assignments from the two samples containing ECF derivatized amino-acids, as we can limit the assignments to those that match expected derivatizations of the known amino-acids (see Supplemental for the expected EMFs, and expected relative NAPs for the individual IMFs).

@@

For large datasets, we expect that more correct peak intensities will result in better agreement between sample normalized peak intensities within a sample class. 
If incorrect peak intensities are being reported, then we would expect between class to within class variances to be very random, resulting in F-statistics close to 1.
